### 2013-02-24 article.tex
Add article text
Signed-off-by: Michael Barton <mail@michaelbarton.me.uk>

% Template for PLoS
% Version 1.0 January 2009
%
\documentclass[10pt]{article}

% cite package, to clean up citations in the main text. Do not remove.
\usepackage{cite}

\usepackage{color}

% Use doublespacing - comment out for single spacing
%\usepackage{setspace} 
%\doublespacing

% Text layout
\topmargin 0.0cm
\oddsidemargin 0.5cm
\evensidemargin 0.5cm
\textwidth 16cm 
\textheight 21cm

% Bold the 'Figure #' in the caption and separate it with a period
% Captions will be left justified
\usepackage[labelfont=bf,labelsep=period,justification=raggedright]{caption}

% Use the PLoS provided bibtex style
\bibliographystyle{article}

% Remove brackets from numbering in List of References
\makeatletter
\renewcommand{\@biblabel}[1]{\quad#1.}
\makeatother

% Leave date blank
\date{}

\pagestyle{myheadings}
%% ** EDIT HERE **

%% ** EDIT HERE **
%% PLEASE INCLUDE ALL MACROS BELOW

%% END MACROS SECTION

\begin{document}

% Title must be 150 characters or less
\begin{flushleft}
{\Large
\textbf{}
}
% Insert Author names, affiliations and corresponding author email.
\\
Michael D. Barton$^{1\ast}$, 
Hazel A. Barton$^{1}$
\\
\bf{1} Biology Department, The University of Akron, Akron, OH, 44325-3908, USA
\\
$\ast$ E-mail: mail@michaelbarton.me.uk
\end{flushleft}

\section*{Abstract}

The increasing accessibility and reduced costs of sequencing has made genome
analysis accessible to more and more researchers. In contrast there remains a
steep learning curve in the subsequent computational steps to required to
process raw sequence reads into a database-deposited genome sequence. Here we
describe ``Genomer,'' a tool that simplifies the manual tasks of finishing and
uploading a genome sequence to a database. Genomer can format a genome scaffold
into the common files required to submit to GenBank. Genomer simplifies editing
a genome scaffold by allowing an human-readable YAML format file to be edited
instead of large sequence files. Genomer is written as a command line tool and
is an effort to make the manual, and increasingly common, process of genome
scaffolding more robust and reproducible. Extensive documentation and video
tutorials are available at http://next.gs.

\section*{Introduction}

The decreasing costs and increasing diversity of high-throughput sequencing
methods is making genome analysis a common method to tackle outstanding
questions in microbiology \cite{loman2012b}. Sequencing can produce
$10e^{6}$-$10e^{10}$ short nucleotide reads, which must be assembled into
larger contigs then scaffolded into larger megabase fragments or complete
chromosomes. These larger DNA sequences are then annotated with genomic
features such as protein-coding genes. Each of these steps can provide
imperfect results and remains the subject of active research
\cite{earl2011,quail2012,beckloff2012}, and manual curation by a researcher may
still improve upon these automated approaches.

Manually editing a genome sequence or annotation is a non-trivial problem
requiring significant effort on a researcher's part. The large size of FASTA
sequence files or general feature format (GFF) annotation files makes changing
these files in a text editor difficult. Tasks such as organising project files,
adding additional sequences to an assembly, updating incorrect annotations or
preparing files for database deposition all divert effort from the main goal of
analysing a genome. As a result genome projects have a steep learning curve for
researchers embarking on genome sequencing for the first time, while still
requiring effort for experienced researchers.

We anticipate these problems have resulted in individual groups independently
producing in-house programming scripts to automate repetitive tasks. An example
of such tasks include finding protein names that do not match a convention of
three lower case characters followed by one upper case character (abcD). This
is manually laborious but simple to automate computationally. Individual
research groups producing their own scripts for such tasks however leads to a
repetition of effort. In contrast open-source projects can lead to a pooling of
community efforts resulting in a higher standard of software and preventing
\cite{ince2012}.

During the process of sequencing several \emph{Pseudomonas fluorescens} strains
we developed a tool, called ``Genomer'', to ease and simplify many of the
common steps in a genome project. Several graphical user interface tools are
available to work with genome projects \cite{tanaka2006, wilkinson2002,
lopez2011, carver2012, gordon2003}. Genomer instead provides a simple, command
line interface for genome finishing allowing scriptable automation and
reproducible results. We have open-sourced this tool under and MIT license and
here we describe its application for simplifying steps in a microbial genome
project.

\section*{Implementation}

Genomer is written in the Ruby programming language \cite{ruby-lang,goto2010}
and tested against the 1.8.7 and 1.9.3 versions. Genomer is built upon the
Rubygems and Bundler Ruby package management libraries. Rubygems is a software
management system for Ruby which allows automatic downloading and installation
of third-party software libraries from the rubygems.org website. Bundler is
software for determining which specific versions of the available third
libraries should be used.

Genomer uses these two libraries to create a plugin system where third-party
software can be written and included. Plugin creators need only to prefix the
name of their plugin with `genomer-plugin-' and upload this to rubygems.org or
make the plugin available as a public git repository. A plugin is specified
within the project in a file named `Gemfile.' The `bundle update` command will
then automatically download and install the required plugins.

Genomer is implemented as a command line tool and tested on both Mac OS X and
Linux systems. The latest version of genomer can installed from the Rubygems
package management system with the command `gem install genomer`. Genomer is
then invoked on the command line with the command `genomer`. Genomer provides
extensive documentation via man pages and on the genomer website at
http://next.gs.

\section*{Results and Discussion}

Genomer was developed during our \emph{P. fluorescens} genome projects
\cite{barton2012} to automate and simplify the manual steps required when
finishing a microbial genome. Genomer includes functions to simplify the
following: updating and modifying a genome sequence, editing genome
annotations, and preparing the required files to submit a draft or complete
genome to a sequence database such as GenBank. Genomer is also designed to
allow researchers to add their own functionality to a genome project through
third-party plugins.

This software was written for the command-line to use in shells such as bash or
zsh. Command line tools are simple to automate using scripting, allowing
reproducible genome finishing and sharing genome projects between researchers.
For instance, during our microbial genome project we automated Genomer using
GNU Make build files, allowing the finishing steps to be repeated automatically
using the `make` command. Additionally we stored each version of the GNU Make
build file with git \cite{git-scm} to track changes to the project and to allow
errors to be reverted from earlier versions. Example build files from our
genome projects can be found on GitHub for a simple plasmid
\cite{plasmid-github} and more complex \cite{genome-github} sequence.

We divided this tool into modular parts each called a `plugin.' Genomer
provides no functionality itself, but rather serves to make the functionality
of plugins available for use at the command line. For example to validate the
correctness of genome annotations, the genomer-plugin-validate plugin should be
installed which then allows the required functionality to be through the
`genomer` command. Presently, we have developed and released three plugins:
viewing a genome sequence in common bioinformatics file formats, validating the
correctness of genome annotations for common errors, and provide summaries of a
genome sequence. Each of these plugins are available on GitHub and their use
described is in the documentation at http://next.gs.

The Genomer plugin system was built using the existing Ruby package management
system: RubyGems and Bundler. This allows plugins to be automatically installed
without requiring manual downloading and compilation by the user. We anticipate
this will eliminate a common problem where software may require technical
expertise to first compile and install before being used. A plugin system also
allows each plugin to be developed on different release schedules without
requiring Genomer itself to be updated, thereby providing a stable interface.
Each Genomer plugin can be locked to specific versions to prevent changes in
functionality between different plugin versions. This addresses situations
whereby backwards incompatible software changes break an existing genomics when
a recent version is available.

We have released Genomer as open source on GitHub \cite{genomer-github} and
have prepared documentation and video tutorials at http://next.gs. As we have
used this tool extensively to simplify our own genome projects we believe
genomer may also be useful to others in the field. A practice described as
``eating your own dog food'' \cite{harrison2006}. We will illustrate the
possible usefulness of genomer with two use cases: updating the annotations of
a genome scaffold during the finishing process, and preparing genome files for
submission to Genbank.

\subsection*{First Use Case: Continuous Genome Finishing}

Genome annotation is required to identify the features responsible for
phenotype. Online tools such as the Integrated Microbial Genomes system
\cite{markowitz2006} provide an online service whereby a nucleotide sequence
may be submitted and a subsequent file of predicted genomic features returned.
These annotations are required when submitting a genome sequence to a database.
Once annotated modifying the sequence becomes problematic as this requires
corresponding changes in all downstream annotated coordinates. This therefore
presents a problem whereby either no more updates can be made to the sequence
or the delta difference in annotation coordinates must be propagated to all
downstream locations. The former option is often choosen as being simpler and
therefore the genome sequence must be considered `frozen' once annotated and no
more changes can be made.

During our own \emph{P. fluorescens} genome sequencing projects, we aimed to
continuously improve the quality of the genome sequence by closing gaps using
PCR and Sanger sequencing while at the same time analysing the observed genomic
features. We therefore incorporated functionality into Genomer to allow
annotation coordinates to be automatically updated following such changes to
the sequence. Genomer allows this by organising a genome project as follows:

\begin{enumerate}

\item{Sequencing reads are assembled into contigs with a high degree of
confidence. This is done by using various assembly software and testing
different software parameters \cite{earl2011}. The generated contigs are then
compared between assembly replicates using nucmer from the Mummer software
package \cite{kurtz2004}. The FASTA file of contig sequences are then added to
the Genomer project.}

\item{The assembled contigs are then annotated and the GFF3 file of contig
annotations is added to the Genomer project.}

\item{A genome scaffold is then created using the scaffolder file format
\cite{barton2012}. This specifies the order and orientation of contigs in the
scaffold allowing any unresolved gap regions to also be specified. Paired-end
sequencing, comparison to reference genomes, or PCR amplification can be used
to estimate the positions of observed contigs in the true genome sequence.}

\item{The genomer-plugin-view plugin is then added to the Gemfile in the root
of the project directory. The command `bundle update`, run in the project root,
will then download and install this plugin making it available for use in the
Genomer project.}

\item{Different views can then be generated from the scaffold using the
`genomer view` command. For instance, a GFF3 file of the scaffolded annotations
may be generated. Should the originating contig annotations or the genome
scaffold be changed then rerunning the `genomer view` command will
correspondingly update the annotation coordinates for the scaffold.}

\end{enumerate}

This allows the scaffold file to be continuously improved even when already
annotated. For instance, additional PCR sequences can be used to close gaps and
improve the genome sequence, without requiring reannotation or manually
updating all annotation coordinates.

\subsection*{Second Use Case: Preparing files for submission to Genbank}

Generating a usable genome sequence requires sequencing, assembly of reads, and
annotation. A further step is also required: preparation of genomic files for
deposition in a database. An example is the Genbank database which requires the
genome annotations to be prepared in a custom table format. Additional files
are also required when the genome assembly is incomplete: a FASTA file
containing the individual contig sequences and `a golden path' (AGP)
\cite{agp-spec} file which describes the placement and direction of constituent
contigs in the draft genome sequence. The requirements to generate these
specific files can be time-consuming to complete and difficult to understand
for researchers new to field. We therefore Genomer during our own genome
projects to simplify and automate this process.

The genomer-view-plugin can be used to generate for submission to GenBank:
assembled FASTA, contig FASTA, annotation tables, and scaffold layout in AGP.
This includes FASTA sequences for both the scaffolded and individual contigs, a
GenBank table file of contig annotations, and an AGP file for the order of
contigs in the scaffold. Additional options may also be used to add a prefix to
gene annotation IDs, or begin locus ID numbering at the origin of the sequence,
thereby simplifying the process of creating the required files for submission.

Genomer does not produce the .asn which must be submitted to GenBank, instead
the generated files are used as the input for the tbl2asn or sequin tools. We
emphasise that the purpose of Genomer is not to reinvent the functionality
already provided by current NCBI resources, but instead to simplify the process
of producing the required inputs. Our example \emph{P. fluorescens} GNU Make
file [genome][] illustrates how Genomer may be combined with tbl2asn to produce
the required .asn file for submission.

\subsection*{Summary}

Genomer is an open-source tool to simplify and automate repetitive and
time-consuming tasks required when finishing a microbial genome project. This
software is available on GitHub \cite{genomer-github} with documentation and
video tutorials on the http://next.gs website. This tool has been useful in our
own genome projects and we believe it will also be useful to other researchers
especially smaller research groups entering the field for the first time.

% Do NOT remove this, even if you are not including acknowledgments
\section*{Acknowledgments}

This work was supported by the National Institute for Health: IDeA Network of
Biomedical Research Excellence (KY-INBRE) grant (NIH 2P20 RR016481-09) and the
NIH R15 AREA Program grant (R15GM079775).

%\section*{References}
% The bibtex filename
\bibliography{article}

\end{document}

### 2013-02-24 article.tex
Format inline code
Signed-off-by: Michael Barton <mail@michaelbarton.me.uk>

@@
 software can be written and included. Plugin creators need only to prefix the
-name of their plugin with `genomer-plugin-' and upload this to rubygems.org or
-make the plugin available as a public git repository. A plugin is specified
-within the project in a file named `Gemfile.' The `bundle update` command will
-then automatically download and install the required plugins.
+name of their plugin with \verb+genomer-plugin-+ and upload this to
+rubygems.org or make the plugin available as a public git repository. A plugin
+is specified within the project in a file named `Gemfile.' The \verb+bundle update+
+command will then automatically download and install the required plugins.
 
@@
 Linux systems. The latest version of genomer can installed from the Rubygems
-package management system with the command `gem install genomer`. Genomer is
-then invoked on the command line with the command `genomer`. Genomer provides
-extensive documentation via man pages and on the genomer website at
+package management system with the command \verb+gem install genomer+. Genomer
+is then invoked on the command line with the command \verb+genomer+. Genomer
+provides extensive documentation via man pages and on the genomer website at
 http://next.gs.
@@
 GNU Make build files, allowing the finishing steps to be repeated automatically
-using the `make` command. Additionally we stored each version of the GNU Make
-build file with git \cite{git-scm} to track changes to the project and to allow
-errors to be reverted from earlier versions. Example build files from our
+using the \verb+make+ command. Additionally we stored each version of the GNU
+Make build file with git \cite{git-scm} to track changes to the project and to
+allow errors to be reverted from earlier versions. Example build files from our
 genome projects can be found on GitHub for a simple plasmid
@@
 installed which then allows the required functionality to be through the
-`genomer` command. Presently, we have developed and released three plugins:
-viewing a genome sequence in common bioinformatics file formats, validating the
-correctness of genome annotations for common errors, and provide summaries of a
-genome sequence. Each of these plugins are available on GitHub and their use
-described is in the documentation at http://next.gs.
+\verb+genomer+ command. Presently, we have developed and released three
+plugins: viewing a genome sequence in common bioinformatics file formats,
+validating the correctness of genome annotations for common errors, and provide
+summaries of a genome sequence. Each of these plugins are available on GitHub
+and their use described is in the documentation at http://next.gs.
 
@@
 \item{The genomer-plugin-view plugin is then added to the Gemfile in the root
-of the project directory. The command `bundle update`, run in the project root,
-will then download and install this plugin making it available for use in the
-Genomer project.}
+of the project directory. The command \verb+bundle update+, run in the project
+root, will then download and install this plugin making it available for use in
+the Genomer project.}
 
 \item{Different views can then be generated from the scaffold using the
-`genomer view` command. For instance, a GFF3 file of the scaffolded annotations
-may be generated. Should the originating contig annotations or the genome
-scaffold be changed then rerunning the `genomer view` command will
+\verb+genomer view+ command. For instance, a GFF3 file of the scaffolded
+annotations may be generated. Should the originating contig annotations or the
+genome scaffold be changed then rerunning the \verb+genomer view+ command will
 correspondingly update the annotation coordinates for the scaffold.}

### 2013-02-24 article.tex
Spell check
Signed-off-by: Michael Barton <mail@michaelbarton.me.uk>

@@
 a genome scaffold during the finishing process, and preparing genome files for
-submission to Genbank.
+submission to GenBank.
 
@@
 or the delta difference in annotation coordinates must be propagated to all
-downstream locations. The former option is often choosen as being simpler and
+downstream locations. The former option is often chosen as being simpler and
 therefore the genome sequence must be considered `frozen' once annotated and no
@@
 
-\subsection*{Second Use Case: Preparing files for submission to Genbank}
+\subsection*{Second Use Case: Preparing files for submission to GenBank}
 
@@
 annotation. A further step is also required: preparation of genomic files for
-deposition in a database. An example is the Genbank database which requires the
+deposition in a database. An example is the GenBank database which requires the
 genome annotations to be prepared in a custom table format. Additional files
@@
 % Do NOT remove this, even if you are not including acknowledgments
-\section*{Acknowledgments}
+\section*{Acknowledgements}
 

### 2013-03-01 article.tex

@@
 into the common files required to submit to GenBank. Genomer simplifies editing
-a genome scaffold by allowing an human-readable YAML format file to be edited
+a genome scaffold by allowing a human-readable YAML format file to be edited
 instead of large sequence files. Genomer is written as a command line tool and
@@
 sequence files or general feature format (GFF) annotation files makes changing
-these files in a text editor difficult. Tasks such as organising project files,
+these in a text editor difficult. Tasks such as organising project files,
 adding additional sequences to an assembly, updating incorrect annotations or
@@
 of such tasks include finding protein names that do not match a convention of
-three lower case characters followed by one upper case character (abcD). This
-is manually laborious but simple to automate computationally. Individual
+three lower case characters followed by one upper case character (abcD). Such
+tasks are manually laborious but simple to automate computationally. Individual
 research groups producing their own scripts for such tasks however leads to a
 repetition of effort. In contrast open-source projects can lead to a pooling of
-community efforts resulting in a higher standard of software and preventing
+community effort resulting in a higher standard of software and preventing
 \cite{ince2012}.
@@
 line interface for genome finishing allowing scriptable automation and
-reproducible results. We have open-sourced this tool under and MIT license and
+reproducible results. We have open-sourced this tool under the MIT license and
 here we describe its application for simplifying steps in a microbial genome
@@
 rubygems.org or make the plugin available as a public git repository. A plugin
-is specified within the project in a file named `Gemfile.' The \verb+bundle update+
-command will then automatically download and install the required plugins.
+is specified within a Genomer project in a file named `Gemfile.' The
+\verb+bundle update+ command will then automatically download and install the
+required plugins.
 
@@
 package management system with the command \verb+gem install genomer+. Genomer
-is then invoked on the command line with the command \verb+genomer+. Genomer
-provides extensive documentation via man pages and on the genomer website at
-http://next.gs.
+is based on our previously described ``scaffolder'' software and is invoked on
+the command line with the command \verb+genomer+. Genomer provides extensive
+documentation via man pages and on the genomer website at http://next.gs.
 
@@
 
-Genomer was developed during our \emph{P. fluorescens} genome projects
-\cite{barton2012} to automate and simplify the manual steps required when
-finishing a microbial genome. Genomer includes functions to simplify the
-following: updating and modifying a genome sequence, editing genome
-annotations, and preparing the required files to submit a draft or complete
-genome to a sequence database such as GenBank. Genomer is also designed to
-allow researchers to add their own functionality to a genome project through
-third-party plugins.
+Genomer was developed during our \emph{P. fluorescens} genome projects to
+automate and simplify the manual steps required when finishing a microbial
+genome. Genomer includes functions to simplify the following: updating and
+modifying a genome sequence, editing genome annotations, and preparing the
+required files to submit a draft or complete genome to a sequence database such
+as GenBank. Genomer is also designed to allow researchers to add their own
+functionality to a genome project through third-party plugins.
 
@@
 genome projects can be found on GitHub for a simple plasmid
-\cite{plasmid-github} and more complex \cite{genome-github} sequence.
+\cite{plasmid-github} and more complex genome \cite{genome-github} sequence.
 
@@
 functionality between different plugin versions. This addresses situations
-whereby backwards incompatible software changes break an existing genomics when
-a recent version is available.
+whereby backwards incompatible software changes can break an existing project's
+workflow.
 
@@
 used this tool extensively to simplify our own genome projects we believe
-genomer may also be useful to others in the field. A practice described as
-``eating your own dog food'' \cite{harrison2006}. We will illustrate the
+genomer may also be useful to others in the field. We will illustrate the
 possible usefulness of genomer with two use cases: updating the annotations of
@@
 specific files can be time-consuming to complete and difficult to understand
-for researchers new to field. We therefore Genomer during our own genome
+for researchers new to field. We used Genomer extensively during our own genome
 projects to simplify and automate this process.
@@
 
-Genomer does not produce the .asn which must be submitted to GenBank, instead
-the generated files are used as the input for the tbl2asn or sequin tools. We
-emphasise that the purpose of Genomer is not to reinvent the functionality
-already provided by current NCBI resources, but instead to simplify the process
-of producing the required inputs. Our example \emph{P. fluorescens} GNU Make
-file [genome][] illustrates how Genomer may be combined with tbl2asn to produce
-the required .asn file for submission.
+Genomer does not produce the .asn file which must be submitted to GenBank,
+instead the generated files are used as the input for the tbl2asn or sequin
+tools. We emphasise that the purpose of Genomer is not to reinvent the
+functionality already provided by current NCBI resources, but instead to
+simplify the process of producing the required inputs. Our example \emph{P.
+fluorescens} GNU Make file \cite{genome-github} illustrates how Genomer may be
+combined with tbl2asn to produce the required .asn file for submission.
 

### 2013-03-01 article.tex
Add title
Signed-off-by: Michael Barton <mail@michaelbarton.me.uk>
Further revisions for clarity
Signed-off-by: Michael Barton <mail@michaelbarton.me.uk>

@@
 {\Large
-\textbf{}
+\textbf{Genomer - a Swiss army knife for genome scaffolding.}
 }
@@
 uploading a genome sequence to a database. Genomer can format a genome scaffold
-into the common files required to submit to GenBank. Genomer simplifies editing
-a genome scaffold by allowing a human-readable YAML format file to be edited
-instead of large sequence files. Genomer is written as a command line tool and
-is an effort to make the manual, and increasingly common, process of genome
-scaffolding more robust and reproducible. Extensive documentation and video
-tutorials are available at http://next.gs.
+into the common files required for submission to GenBank. Genomer simplifies
+editing a genome scaffold by allowing a human-readable YAML format file to be
+edited instead of large sequence files. Genomer is written as a command line
+tool and is an effort to make the manual, and increasingly common, process of
+genome scaffolding more robust and reproducible. Extensive documentation and
+video tutorials are available at http://next.gs.
 
@@
 The decreasing costs and increasing diversity of high-throughput sequencing
-methods is making genome analysis a common method to tackle outstanding
+methods is making genome analysis a common method to tackle unresolved
 questions in microbiology \cite{loman2012b}. Sequencing can produce
@@
 chromosomes. These larger DNA sequences are then annotated with genomic
-features such as protein-coding genes. Each of these steps can provide
-imperfect results and remains the subject of active research
-\cite{earl2011,quail2012,beckloff2012}, and manual curation by a researcher may
-still improve upon these automated approaches.
+features such as protein-coding genes. Each of these required steps can however
+produce imperfect results and remains the subject of active research
+\cite{earl2011,quail2012,beckloff2012}. Therefore manual curation of a genome
+sequence and corresponding annotations may still improve upon the results of
+automated methods.
 
@@
 community effort resulting in a higher standard of software and preventing
-\cite{ince2012}.
+reinvention of the same code \cite{ince2012}.
 
@@
 we developed a tool, called ``Genomer'', to ease and simplify many of the
-common steps in a genome project. Several graphical user interface tools are
+common steps in our genome project. Several graphical user interface tools are
 available to work with genome projects \cite{tanaka2006, wilkinson2002,
@@
 the command line with the command \verb+genomer+. Genomer provides extensive
-documentation via man pages and on the genomer website at http://next.gs.
+documentation via UNIX manual pages and on the genomer website at
+http://next.gs.
 
@@
 genome projects can be found on GitHub for a simple plasmid
-\cite{plasmid-github} and more complex genome \cite{genome-github} sequence.
+\cite{plasmid-github} and a more complex genome \cite{genome-github} sequence.
 

### 2013-04-17 article.tex
Make features clear at start of results
Signed-off-by: Michael Barton <mail@michaelbarton.me.uk>

@@
 automate and simplify the manual steps required when finishing a microbial
-genome. Genomer includes functions to simplify the following: updating and
-modifying a genome sequence, editing genome annotations, and preparing the
-required files to submit a draft or complete genome to a sequence database such
-as GenBank. Genomer is also designed to allow researchers to add their own
-functionality to a genome project through third-party plugins.
+genome. Genomer provides the following functionality to facilitate this:
+
+\begin{description}
+
+  \item[Simple editting of a draft genome sequence.]{Genomer is built on the
+  existing Scaffolder \cite{barton2012} file format for assembling a draft
+  genome sequence. This YAML \cite{yaml} format simplifies the process of
+  re-organising and trimming contigs in a draft genome without having to edit
+  large nucleotide sequences manually.}
+
+  \item[Simple mapping of annotations to the generated draft sequence.]{Genomer
+  maps the coordinates of annotations on individual to repsective position in
+  the draft genome sequence. This allows the scaffold file to be continuously
+  improved even when already annotated. For instance, additional PCR sequences
+  can be used to close gaps and improve the genome sequence, without requiring
+  reannotation or manually updating all annotation coordinates.}
+
+  \item[Generation of files for submission to GenBank.]{Submitting a genome
+  sequence to GenBank database requires several specific files to be prepared.
+  These may include a FASTA file of the draft sequence, a table of annotations,
+  a FASTA file containing the individual contig sequences, and `a golden path'
+  (AGP) \cite{agp-spec} file describing the placement of contigs. Genomer
+  automates the generation of all these files.}
+
+  \item[Integration with the command line.]{Genomer is built as a command line
+  tool built around plain text files following the Linux philosophy. This
+  allows for integration with tools such as GNU Make or git, and also to allow
+  reproducible scripting of the process of finishing a genome.}
+
+\end{description}
 
@@
 
-\item{Sequencing reads are assembled into contigs with a high degree of
-confidence. This is done by using various assembly software and testing
-different software parameters \cite{earl2011}. The generated contigs are then
-compared between assembly replicates using nucmer from the Mummer software
-package \cite{kurtz2004}. The FASTA file of contig sequences are then added to
-the Genomer project.}
-
-\item{The assembled contigs are then annotated and the GFF3 file of contig
-annotations is added to the Genomer project.}
+\item{A genomer project is organised around a set of already assembled and
+annotated contigs. For example we assemble our contigs using the A5 pipeline
+\cite{tritt2012} and and then subsequently annotate them using the Integrated
+Microbial Genomes resource \cite{markowitz2006}. We then use these contigs as a
+starting point to build a draft genome sequence in genomer.}
 
@@
 
-This allows the scaffold file to be continuously improved even when already
-annotated. For instance, additional PCR sequences can be used to close gaps and
-improve the genome sequence, without requiring reannotation or manually
-updating all annotation coordinates.
-
 \subsection*{Second Use Case: Preparing files for submission to GenBank}
-
-Generating a usable genome sequence requires sequencing, assembly of reads, and
-annotation. A further step is also required: preparation of genomic files for
-deposition in a database. An example is the GenBank database which requires the
-genome annotations to be prepared in a custom table format. Additional files
-are also required when the genome assembly is incomplete: a FASTA file
-containing the individual contig sequences and `a golden path' (AGP)
-\cite{agp-spec} file which describes the placement and direction of constituent
-contigs in the draft genome sequence. The requirements to generate these
-specific files can be time-consuming to complete and difficult to understand
-for researchers new to field. We used Genomer extensively during our own genome
-projects to simplify and automate this process.
 

### 2013-04-18 article.tex
Add final reviewer requested revisions
Signed-off-by: Michael Barton <mail@michaelbarton.me.uk>

@@
 analysis accessible to more and more researchers. In contrast there remains a
-steep learning curve in the subsequent computational steps to required to
-process raw sequence reads into a database-deposited genome sequence. Here we
-describe ``Genomer,'' a tool that simplifies the manual tasks of finishing and
-uploading a genome sequence to a database. Genomer can format a genome scaffold
-into the common files required for submission to GenBank. Genomer simplifies
-editing a genome scaffold by allowing a human-readable YAML format file to be
-edited instead of large sequence files. Genomer is written as a command line
-tool and is an effort to make the manual, and increasingly common, process of
-genome scaffolding more robust and reproducible. Extensive documentation and
-video tutorials are available at http://next.gs.
+steep learning curve in the subsequent computational steps required to process
+raw sequence reads into a database-deposited genome sequence. Here we describe
+``Genomer,'' a tool that simplifies the manual tasks of finishing and uploading
+a genome sequence to a database. Genomer can format a genome scaffold into the
+common files required for submission to GenBank. Genomer simplifies editing a
+genome scaffold by allowing a human-readable YAML format file to be edited
+instead of large sequence files. Genomer is written as a command line tool and
+is an effort to make the manual, and increasingly common, process of genome
+scaffolding more robust and reproducible. Extensive documentation and video
+tutorials are available at http://next.gs.
 
@@
 
-  \item[Simple editting of a draft genome sequence.]{Genomer is built on the
-  existing Scaffolder \cite{barton2012} file format for assembling a draft
-  genome sequence. This YAML \cite{yaml} format simplifies the process of
-  re-organising and trimming contigs in a draft genome without having to edit
+  \item[Simple editing of a draft genome sequence.]{Genomer is built on the
+  existing Scaffolder \cite{barton2012} file format for assembling draft genome
+  sequences. This YAML \cite{yaml} format simplifies the process of
+  re-organising and trimming contigs in draft genome without having to edit
   large nucleotide sequences manually.}
 
-  \item[Simple mapping of annotations to the generated draft sequence.]{Genomer
-  maps the coordinates of annotations on individual to repsective position in
-  the draft genome sequence. This allows the scaffold file to be continuously
-  improved even when already annotated. For instance, additional PCR sequences
-  can be used to close gaps and improve the genome sequence, without requiring
-  reannotation or manually updating all annotation coordinates.}
+  \item[Mapping of annotations onto the assembled sequence.]{Genomer maps the
+  coordinates of contig annotations to respective positions in the draft genome
+  sequence. This allows the scaffold file to be continuously improved and
+  updated even after annotation. For instance, additional PCR sequences can be
+  used to close gaps in the assembly without requiring reannotation or manually
+  updating annotation coordinates.}
 
   \item[Generation of files for submission to GenBank.]{Submitting a genome
-  sequence to GenBank database requires several specific files to be prepared.
-  These may include a FASTA file of the draft sequence, a table of annotations,
-  a FASTA file containing the individual contig sequences, and `a golden path'
-  (AGP) \cite{agp-spec} file describing the placement of contigs. Genomer
-  automates the generation of all these files.}
+  sequence to GenBank database requires generating specific files. These may
+  include a FASTA file of the draft sequence, a table of annotations, a FASTA
+  file containing the individual contig sequences, and `a golden path' (AGP)
+  \cite{agp-spec} file describing the placement of contigs. Genomer automates
+  the generation of all these files from the assembled draft genome.}
+
+  \item[A stable interface and streamlined install process]{Genomer provides a
+  plugin system built using the existing Ruby package management system:
+  RubyGems and Bundler. This allows genomer and its plugins to be automatically
+  installed without requiring manual downloading and compilation by the user.
+  This eliminates a common problem in bioinformatics where software may require
+  technical expertise to first compile and install before being used. Plugins
+  can also be locked to specific versions to prevent backwards incompatible
+  software changes breaking an existing workflow.}
 
   \item[Integration with the command line.]{Genomer is built as a command line
-  tool built around plain text files following the Linux philosophy. This
-  allows for integration with tools such as GNU Make or git, and also to allow
-  reproducible scripting of the process of finishing a genome.}
+  tool built around plain text files. This allows for integration with common
+  Linux tools tools such as GNU Make or git. This thereby allow reproducible
+  scripting of the genome finishing process.}
 
@@
 
-This software was written for the command-line to use in shells such as bash or
-zsh. Command line tools are simple to automate using scripting, allowing
+As Genomer is written the command-line to use in shells such as bash or zsh.
+Command line tools are simple to automate using scripting, allowing
 reproducible genome finishing and sharing genome projects between researchers.
@@
 GNU Make build files, allowing the finishing steps to be repeated automatically
-using the \verb+make+ command. Additionally we stored each version of the GNU
-Make build file with git \cite{git-scm} to track changes to the project and to
-allow errors to be reverted from earlier versions. Example build files from our
-genome projects can be found on GitHub for a simple plasmid
-\cite{plasmid-github} and a more complex genome \cite{genome-github} sequence.
+using the \verb+make+ command. This is an advantage of command-line tools
+versus GUI-based tools, the latter are harder to automate and share steps taken
+to produce results. Therefore Genomer may be of particular interest to
+bioinformaticians who prefer automated approaches and scripting on the command
+line.
 
-We divided this tool into modular parts each called a `plugin.' Genomer
-provides no functionality itself, but rather serves to make the functionality
-of plugins available for use at the command line. For example to validate the
-correctness of genome annotations, the genomer-plugin-validate plugin should be
-installed which then allows the required functionality to be through the
-\verb+genomer+ command. Presently, we have developed and released three
-plugins: viewing a genome sequence in common bioinformatics file formats,
-validating the correctness of genome annotations for common errors, and provide
-summaries of a genome sequence. Each of these plugins are available on GitHub
-and their use described is in the documentation at http://next.gs.
-
-The Genomer plugin system was built using the existing Ruby package management
-system: RubyGems and Bundler. This allows plugins to be automatically installed
-without requiring manual downloading and compilation by the user. We anticipate
-this will eliminate a common problem where software may require technical
-expertise to first compile and install before being used. A plugin system also
-allows each plugin to be developed on different release schedules without
-requiring Genomer itself to be updated, thereby providing a stable interface.
-Each Genomer plugin can be locked to specific versions to prevent changes in
-functionality between different plugin versions. This addresses situations
-whereby backwards incompatible software changes can break an existing project's
-workflow.
+Additionally as Genomer uses plain text it is easy to store versions the
+project with revision control such as git \cite{git-scm}. This allows tracking
+changes to the project and to reverting errors back to earlier versions.
+Example build files from our genome projects can be found on GitHub for a
+simple plasmid \cite{plasmid-github} and a more complex genome
+\cite{genome-github} sequence.
 
@@
 
-\subsection*{First Use Case: Continuous Genome Finishing}
+\subsection*{Example use case with Genomer}
 
-Genome annotation is required to identify the features responsible for
-phenotype. Online tools such as the Integrated Microbial Genomes system
+Online tools such as the Integrated Microbial Genomes system
 \cite{markowitz2006} provide an online service whereby a nucleotide sequence
 may be submitted and a subsequent file of predicted genomic features returned.
-These annotations are required when submitting a genome sequence to a database.
 Once annotated modifying the sequence becomes problematic as this requires
@@
 or the delta difference in annotation coordinates must be propagated to all
-downstream locations. The former option is often chosen as being simpler and
-therefore the genome sequence must be considered `frozen' once annotated and no
-more changes can be made.
+downstream locations.
 
@@
 annotation coordinates to be automatically updated following such changes to
-the sequence. Genomer allows this by organising a genome project as follows:
+the sequence.
 
@@
 
-\item{A genomer project is organised around a set of already assembled and
-annotated contigs. For example we assemble our contigs using the A5 pipeline
-\cite{tritt2012} and and then subsequently annotate them using the Integrated
-Microbial Genomes resource \cite{markowitz2006}. We then use these contigs as a
-starting point to build a draft genome sequence in genomer.}
+  \item{A Genomer project is organised around a set of already assembled and
+  annotated contigs. For example we assemble our reads into contigs using the
+  A5 pipeline \cite{tritt2012} and and then subsequently annotate them using
+  the Integrated Microbial Genomes resource \cite{markowitz2006}. We then use
+  these contigs as a starting point to build a draft genome sequence in
+  Genomer.}
 
-\item{A genome scaffold is then created using the scaffolder file format
+  \item{A genome scaffold is written in the scaffolder file format
   \cite{barton2012}. This specifies the order and orientation of contigs in the
-scaffold allowing any unresolved gap regions to also be specified. Paired-end
-sequencing, comparison to reference genomes, or PCR amplification can be used
-to estimate the positions of observed contigs in the true genome sequence.}
+  scaffold. Any unresolved gap regions can also be specified in this file.
+  Paired-end sequencing or comparison to reference genomes are a source of
+  contig orientation and order. In our genome projects we determine contig
+  order by aligning contigs to a reference genome using nucmer
+  \cite{kurtz2004}.}
 
-\item{The genomer-plugin-view plugin is then added to the Gemfile in the root
-of the project directory. The command \verb+bundle update+, run in the project
-root, will then download and install this plugin making it available for use in
-the Genomer project.}
+  \item{The \verb+genomer view+ command can be used to generate files of the
+  assembled scaffold. These may include the FASTA file of the assembly or a
+  GFF3 file of the annotations in the assembled genome. These files can then be
+  used for downstream analysis of the sequence.}
 
-\item{Different views can then be generated from the scaffold using the
-\verb+genomer view+ command. For instance, a GFF3 file of the scaffolded
-annotations may be generated. Should the originating contig annotations or the
-genome scaffold be changed then rerunning the \verb+genomer view+ command will
-correspondingly update the annotation coordinates for the scaffold.}
+  \item{The genome scaffold can be continuously updated through closing any
+  remaining gaps. During our microbe sequencing projects we closed gaps through
+  a combination of Sanger PCR and further detailed \emph{in silico} analysis.
+  Any changes to the scaffold file automatically propagate to generated files
+  when the \verb+genomer view+ command is rerun.}
+
+  \item{When researchers are satisfied with the status of their draft the
+  required files for submission to GenBank can be generated: assembled FASTA,
+  contig FASTA, annotation tables, and scaffold layout in AGP. Additional
+  options are used to add the required prefix to gene annotation IDs and begin
+  locus ID numbering at the origin of the sequence. Genomer does not produce
+  the .asn file which must be submitted to GenBank, but instead the files for
+  the input to tbl2asn or sequin. }
 
@@
 
-\subsection*{Second Use Case: Preparing files for submission to GenBank}
+\subsection*{Limitations}
 
-The genomer-view-plugin can be used to generate for submission to GenBank:
-assembled FASTA, contig FASTA, annotation tables, and scaffold layout in AGP.
-This includes FASTA sequences for both the scaffolded and individual contigs, a
-GenBank table file of contig annotations, and an AGP file for the order of
-contigs in the scaffold. Additional options may also be used to add a prefix to
-gene annotation IDs, or begin locus ID numbering at the origin of the sequence,
-thereby simplifying the process of creating the required files for submission.
+Genomer was written to satisfy our needs for building and finishing a draft
+microbial genomes less than 10MBp in size. Draft sequence assembly takes place
+in memory and therefore building very large sequences in memory may limit
+performance. We point out however that sizes of even 3GBp should fit in the
+memory of a modern laptop or desktop computer.
 
-Genomer does not produce the .asn file which must be submitted to GenBank,
-instead the generated files are used as the input for the tbl2asn or sequin
-tools. We emphasise that the purpose of Genomer is not to reinvent the
-functionality already provided by current NCBI resources, but instead to
-simplify the process of producing the required inputs. Our example \emph{P.
-fluorescens} GNU Make file \cite{genome-github} illustrates how Genomer may be
-combined with tbl2asn to produce the required .asn file for submission.
+Genomer simplifies the process of generating the annotation table file required
+for submission to GenBank where only gene entries are required. Specifically a
+user need only provide a GFF3 file of annotations and Genomer can automatically
+generate the additional encoded protein features. This however assumes no
+intron/exon structure and therefore users wishing to submit annotations with
+alternative splicing cannot take advantage of this genomer feature.
 

### 2013-04-18 article.tex
Generate diff of article changes
Signed-off-by: Michael Barton <mail@michaelbarton.me.uk>

@@
   existing Scaffolder \cite{barton2012} file format for assembling draft genome
-  sequences. This YAML \cite{yaml} format simplifies the process of
-  re-organising and trimming contigs in draft genome without having to edit
-  large nucleotide sequences manually.}
+  sequences. This YAML format simplifies the process of re-organising and
+  trimming contigs in draft genome without having to edit large nucleotide
+  sequences manually.}
 

### 2013-04-19 article.tex
Proof read manuscript
Signed-off-by: Michael Barton <mail@michaelbarton.me.uk>

@@
 raw sequence reads into a database-deposited genome sequence. Here we describe
-``Genomer,'' a tool that simplifies the manual tasks of finishing and uploading
-a genome sequence to a database. Genomer can format a genome scaffold into the
-common files required for submission to GenBank. Genomer simplifies editing a
+``Genomer,'' a tool to simplify the manual tasks of finishing and uploading a
+genome sequence to a database. Genomer can format a genome scaffold into the
+common files required for submission to GenBank. Genomer simplifies updating a
 genome scaffold by allowing a human-readable YAML format file to be edited
@@
 Manually editing a genome sequence or annotation is a non-trivial problem
-requiring significant effort on a researcher's part. The large size of FASTA
-sequence files or general feature format (GFF) annotation files makes changing
-these in a text editor difficult. Tasks such as organising project files,
-adding additional sequences to an assembly, updating incorrect annotations or
-preparing files for database deposition all divert effort from the main goal of
-analysing a genome. As a result genome projects have a steep learning curve for
-researchers embarking on genome sequencing for the first time, while still
-requiring effort for experienced researchers.
+requiring effort on a researcher's part. The large size of FASTA sequence files
+or general feature format (GFF) annotation files makes changing these in a text
+editor difficult. Tasks such as organising project files, adding additional
+sequences to an assembly, updating incorrect annotations or preparing files for
+database deposition all divert effort from the main goal of analysing a genome.
+As a result genome projects have a steep learning curve for researchers
+embarking on genome sequencing for the first time, while still requiring effort
+for experienced researchers.
 
@@
 reproducible results. We have open-sourced this tool under the MIT license and
-here we describe its application for simplifying steps in a microbial genome
-project.
+here we describe its application in a microbial genome project.
 
@@
   existing Scaffolder \cite{barton2012} file format for assembling draft genome
-  sequences. This YAML format simplifies the process of re-organising and
-  trimming contigs in draft genome without having to edit large nucleotide
-  sequences manually.}
+  sequences. This is format requires only the order and IDs of each contig be
+  specified. This thereby simplifies the process of re-organising and trimming
+  contigs in draft genome without having to edit large nucleotide sequences
+  manually.}
 
@@
   updated even after annotation. For instance, additional PCR sequences can be
-  used to close gaps in the assembly without requiring reannotation or manually
-  updating annotation coordinates.}
+  used to close gaps in the assembly without reannotation or manually updating
+  annotation coordinates.}
 
@@
   \cite{agp-spec} file describing the placement of contigs. Genomer automates
-  the generation of all these files from the assembled draft genome.}
+  the generation of all these files from the scaffold file and corresponding
+  GFF3 file.}
 
@@
 
-As Genomer is written the command-line to use in shells such as bash or zsh.
+Genomer is written for the command-line to use in shells such as bash or zsh.
 Command line tools are simple to automate using scripting, allowing
@@
 using the \verb+make+ command. This is an advantage of command-line tools
-versus GUI-based tools, the latter are harder to automate and share steps taken
-to produce results. Therefore Genomer may be of particular interest to
-bioinformaticians who prefer automated approaches and scripting on the command
-line.
+versus GUI-based tools, the latter are harder to automate and share steps.
+Therefore Genomer may be of particular interest to bioinformaticians who prefer
+automated approaches and scripting on the command line.
 
-Additionally as Genomer uses plain text it is easy to store versions the
-project with revision control such as git \cite{git-scm}. This allows tracking
-changes to the project and to reverting errors back to earlier versions.
+Additionally as Genomer uses plain text it is easy to store versions of the
+project using a revision control system such as git \cite{git-scm}. This allows
+tracking changes to the project and reverting errors back to earlier versions.
 Example build files from our genome projects can be found on GitHub for a
@@
 genomer may also be useful to others in the field. We will illustrate the
-possible usefulness of genomer with two use cases: updating the annotations of
-a genome scaffold during the finishing process, and preparing genome files for
-submission to GenBank.
+possible application of genomer with an example use case.
 
-\subsection*{Example use case with Genomer}
-
-Online tools such as the Integrated Microbial Genomes system
-\cite{markowitz2006} provide an online service whereby a nucleotide sequence
-may be submitted and a subsequent file of predicted genomic features returned.
-Once annotated modifying the sequence becomes problematic as this requires
-corresponding changes in all downstream annotated coordinates. This therefore
-presents a problem whereby either no more updates can be made to the sequence
-or the delta difference in annotation coordinates must be propagated to all
-downstream locations.
-
-During our own \emph{P. fluorescens} genome sequencing projects, we aimed to
-continuously improve the quality of the genome sequence by closing gaps using
-PCR and Sanger sequencing while at the same time analysing the observed genomic
-features. We therefore incorporated functionality into Genomer to allow
-annotation coordinates to be automatically updated following such changes to
-the sequence.
+\subsection*{Genomer use case}
 
@@
   order by aligning contigs to a reference genome using nucmer
-  \cite{kurtz2004}.}
+  \cite{kurtz2004}. A detailed overview of the scaffolder file format can be
+  found on the genomer website.}
 
@@
   assembled scaffold. These may include the FASTA file of the assembly or a
-  GFF3 file of the annotations in the assembled genome. These files can then be
-  used for downstream analysis of the sequence.}
+  GFF3 file of updated annotations corresponding to the assembled genome
+  sequence. These files can then be used for downstream analyses of the
+  sequence.}
 
@@
   remaining gaps. During our microbe sequencing projects we closed gaps through
-  a combination of Sanger PCR and further detailed \emph{in silico} analysis.
-  Any changes to the scaffold file automatically propagate to generated files
-  when the \verb+genomer view+ command is rerun.}
+  a combination of Sanger PCR and \emph{in silico} analysis. Changes to the
+  scaffold file automatically propagate updates in the generated files when the
+  \verb+genomer view+ command is rerun.}
 
@@
   locus ID numbering at the origin of the sequence. Genomer does not produce
-  the .asn file which must be submitted to GenBank, but instead the files for
-  the input to tbl2asn or sequin. }
+  the .asn file which must be submitted to GenBank, but instead the files that
+  may be used as the input to tbl2asn or sequin. }
 

### 2013-04-19 article.tex
Further proof read manuscript
Signed-off-by: Michael Barton <mail@michaelbarton.me.uk>

@@
 database deposition all divert effort from the main goal of analysing a genome.
-As a result genome projects have a steep learning curve for researchers
-embarking on genome sequencing for the first time, while still requiring effort
-for experienced researchers.
+As a result there is a steep learning curve for researchers embarking on genome
+sequencing for the first time, while still requiring effort for experienced
+researchers.
 
@@
 During the process of sequencing several \emph{Pseudomonas fluorescens} strains
-we developed a tool, called ``Genomer'', to ease and simplify many of the
-common steps in our genome project. Several graphical user interface tools are
-available to work with genome projects \cite{tanaka2006, wilkinson2002,
-lopez2011, carver2012, gordon2003}. Genomer instead provides a simple, command
-line interface for genome finishing allowing scriptable automation and
+we developed a tool, called ``Genomer,'' to ease and simplify many of the
+common steps in our genome project. While several graphical user interface
+tools are available to work with genome projects \cite{tanaka2006,
+wilkinson2002, lopez2011, carver2012, gordon1998}, Genomer instead provides a
+simple command line interface. This allows scriptable automation and
 reproducible results. We have open-sourced this tool under the MIT license and
-here we describe its application in a microbial genome project.
+here we describe its application.
 
@@
 \verb+bundle update+ command will then automatically download and install the
-required plugins.
+required plugins plus any gem dependencies these gems rely on.
 
 Genomer is implemented as a command line tool and tested on both Mac OS X and
-Linux systems. The latest version of genomer can installed from the Rubygems
+Linux systems. The latest version of Genomer can installed from the Rubygems
 package management system with the command \verb+gem install genomer+. Genomer
-is based on our previously described ``scaffolder'' software and is invoked on
-the command line with the command \verb+genomer+. Genomer provides extensive
-documentation via UNIX manual pages and on the genomer website at
-http://next.gs.
+is based on our previously described ``Scaffolder'' software \cite{barton2012}
+and is invoked on the command line with the command \verb+genomer+. Genomer
+provides extensive documentation via UNIX manual pages and on the Genomer
+website at http://next.gs.
 
@@
   \item[Mapping of annotations onto the assembled sequence.]{Genomer maps the
-  coordinates of contig annotations to respective positions in the draft genome
-  sequence. This allows the scaffold file to be continuously improved and
-  updated even after annotation. For instance, additional PCR sequences can be
-  used to close gaps in the assembly without reannotation or manually updating
-  annotation coordinates.}
+  coordinates of contig annotations to their respective positions in the draft
+  genome sequence. This allows the scaffold file to be continuously improved
+  and updated even after the contigs have been annotated. For instance,
+  additional PCR sequences can be used to close gaps in the assembly and the
+  location of all annotations be correspondingly updated.}
 
@@
   the generation of all these files from the scaffold file and corresponding
-  GFF3 file.}
+  GFF3 file and removes the effort to produce these manually.}
 
@@
   plugin system built using the existing Ruby package management system:
-  RubyGems and Bundler. This allows genomer and its plugins to be automatically
+  RubyGems and Bundler. This allows Genomer and its plugins to be automatically
   installed without requiring manual downloading and compilation by the user.
@@
 using the \verb+make+ command. This is an advantage of command-line tools
-versus GUI-based tools, the latter are harder to automate and share steps.
-Therefore Genomer may be of particular interest to bioinformaticians who prefer
-automated approaches and scripting on the command line.
+versus GUI-based tools, is that the former are simpler to automate and share
+steps with other researchers. Therefore Genomer may be of particular interest
+to bioinformaticians who prefer automated approaches and scripting on the
+command line.
 
@@
 tracking changes to the project and reverting errors back to earlier versions.
-Example build files from our genome projects can be found on GitHub for a
-simple plasmid \cite{plasmid-github} and a more complex genome
+Example versioned build files from our genome projects can be found on GitHub
+for a simple plasmid \cite{plasmid-github} and a more complex genome
 \cite{genome-github} sequence.
@@
 used this tool extensively to simplify our own genome projects we believe
-genomer may also be useful to others in the field. We will illustrate the
-possible application of genomer with an example use case.
+Genomer may also be useful to others in the field. We will illustrate the
+possible application of Genomer with an example use case.
 
@@
   \cite{kurtz2004}. A detailed overview of the scaffolder file format can be
-  found on the genomer website.}
+  found on the Genomer website.}
 
@@
   locus ID numbering at the origin of the sequence. Genomer does not produce
-  the .asn file which must be submitted to GenBank, but instead the files that
-  may be used as the input to tbl2asn or sequin. }
+  the .asn file which is submitted to GenBank, but instead the files that may
+  be used as the input to tbl2asn or sequin. }
 
@@
 for submission to GenBank where only gene entries are required. Specifically a
-user need only provide a GFF3 file of annotations and Genomer can automatically
-generate the additional encoded protein features. This however assumes no
-intron/exon structure and therefore users wishing to submit annotations with
-alternative splicing cannot take advantage of this genomer feature.
+user need only provide a GFF3 file of annotations and Genomer will generate the
+additional encoded protein features. This however assumes no intron/exon
+structure and therefore users wishing to submit annotations with alternative
+splicing cannot take advantage of this Genomer feature.
 

### 2013-04-22 article.tex
Further proof reading of manuscript
Signed-off-by: Michael Barton <mail@michaelbarton.me.uk>

@@
 The increasing accessibility and reduced costs of sequencing has made genome
-analysis accessible to more and more researchers. In contrast there remains a
-steep learning curve in the subsequent computational steps required to process
-raw sequence reads into a database-deposited genome sequence. Here we describe
+analysis accessible to more and more researchers. Yet there remains a steep
+learning curve in the subsequent computational steps required to process raw
+sequence reads into a database-deposited genome sequence. Here we describe
 ``Genomer,'' a tool to simplify the manual tasks of finishing and uploading a
 genome sequence to a database. Genomer can format a genome scaffold into the
-common files required for submission to GenBank. Genomer simplifies updating a
-genome scaffold by allowing a human-readable YAML format file to be edited
-instead of large sequence files. Genomer is written as a command line tool and
-is an effort to make the manual, and increasingly common, process of genome
-scaffolding more robust and reproducible. Extensive documentation and video
-tutorials are available at http://next.gs.
+common files required for submission to GenBank. This software also simplifies
+updating a genome scaffold by allowing a human-readable YAML format file to be
+edited instead of large sequence files. Genomer is written as a command line
+tool and is an effort to make the manual, and increasingly common, process of
+genome scaffolding more robust and reproducible. Extensive documentation and
+video tutorials are available at http://next.gs.
 
@@
 $10e^{6}$-$10e^{10}$ short nucleotide reads, which must be assembled into
-larger contigs then scaffolded into larger megabase fragments or complete
-chromosomes. These larger DNA sequences are then annotated with genomic
-features such as protein-coding genes. Each of these required steps can however
-produce imperfect results and remains the subject of active research
+larger contigs which are then scaffolded into larger megabase fragments or
+complete chromosomes. These larger DNA sequences are then annotated with
+genomic features such as protein-coding genes. Each of these required steps can
+however produce imperfect results and remains the subject of active research
 \cite{earl2011,quail2012,beckloff2012}. Therefore manual curation of a genome
@@
 As a result there is a steep learning curve for researchers embarking on genome
-sequencing for the first time, while still requiring effort for experienced
+sequencing for the first time, while still requiring effort from experienced
 researchers.
@@
 common steps in our genome project. While several graphical user interface
-tools are available to work with genome projects \cite{tanaka2006,
-wilkinson2002, lopez2011, carver2012, gordon1998}, Genomer instead provides a
-simple command line interface. This allows scriptable automation and
-reproducible results. We have open-sourced this tool under the MIT license and
-here we describe its application.
+tools are available to work with genome projects, including G-InforBio,
+Genquire, Pile-LineGUI, Artemis and Consed \cite{tanaka2006, wilkinson2002,
+lopez2011, carver2012, gordon1998}. Genomer provides a simple command line
+interface which allows scriptable automation and reproducible results. We made
+this software open-source under the MIT license and here we describe its
+application.
 
@@
 
-Genomer is written in the Ruby programming language \cite{ruby-lang,goto2010}
-and tested against the 1.8.7 and 1.9.3 versions. Genomer is built upon the
-Rubygems and Bundler Ruby package management libraries. Rubygems is a software
-management system for Ruby which allows automatic downloading and installation
-of third-party software libraries from the rubygems.org website. Bundler is
-software for determining which specific versions of the available third
-libraries should be used.
+Genomer is written in the Ruby programming language (1.8.7 and 1.9.3)
+\cite{ruby-lang,goto2010} and built upon the Rubygems and Bundler Ruby package
+management libraries. Rubygems is a software management system for Ruby which
+allows automatic downloading and installation of third-party software libraries
+from the rubygems.org website. Bundler is software for determining which
+specific versions of the available third libraries should be used.
 
@@
   This eliminates a common problem in bioinformatics where software may require
-  technical expertise to first compile and install before being used. Plugins
-  can also be locked to specific versions to prevent backwards incompatible
-  software changes breaking an existing workflow.}
+  technical expertise to first compile and install before being used. Install
+  plugins may also be locked to specific versions to prevent backwards
+  incompatible software changes breaking an existing workflow.}
 
   \item[Integration with the command line.]{Genomer is built as a command line
-  tool built around plain text files. This allows for integration with common
-  Linux tools tools such as GNU Make or git. This thereby allow reproducible
-  scripting of the genome finishing process.}
+  tool around plain text files. This allows for integration with common Linux
+  tools tools such as GNU Make or git thereby allowing reproducible scripting
+  of the genome finishing process.}
 
@@
 
-Genomer is written for the command-line to use in shells such as bash or zsh.
-Command line tools are simple to automate using scripting, allowing
-reproducible genome finishing and sharing genome projects between researchers.
-For instance, during our microbial genome project we automated Genomer using
-GNU Make build files, allowing the finishing steps to be repeated automatically
-using the \verb+make+ command. This is an advantage of command-line tools
-versus GUI-based tools, is that the former are simpler to automate and share
-steps with other researchers. Therefore Genomer may be of particular interest
-to bioinformaticians who prefer automated approaches and scripting on the
-command line.
+Genomer is written as a command-line tool for use in shells such as bash or
+zsh. The advantage of command-line tools over GUI-based tools, is that the former
+are simpler to automate and allow the sharing of steps with other researchers
+working on related projects. For instance, during our microbial genome project
+we automated Genomer using GNU Make build files, allowing the finishing steps
+to be repeated automatically using the \verb+make+ command. Genomer may
+therefore be of particular interest to bioinformaticians coordinating in groups
+on the same genome project and those who prefer automated approaches and
+scripting on the command line.
 
-Additionally as Genomer uses plain text it is easy to store versions of the
-project using a revision control system such as git \cite{git-scm}. This allows
-tracking changes to the project and reverting errors back to earlier versions.
-Example versioned build files from our genome projects can be found on GitHub
-for a simple plasmid \cite{plasmid-github} and a more complex genome
-\cite{genome-github} sequence.
+As Genomer uses plain text files to manage the draft genome sequence it is easy
+to store versions of the project using a revision control system such as git
+\cite{git-scm}. This allows tracking changes to the project and reverting
+errors back to earlier versions. Example versioned build files from our genome
+projects can be found on GitHub for a simple plasmid \cite{plasmid-github} and
+a more complex genome \cite{genome-github} sequence.
 
-We have released Genomer as open source on GitHub \cite{genomer-github} and
-have prepared documentation and video tutorials at http://next.gs. As we have
-used this tool extensively to simplify our own genome projects we believe
-Genomer may also be useful to others in the field. We will illustrate the
-possible application of Genomer with an example use case.
+We have released Genomer as open source software on GitHub
+\cite{genomer-github} and have prepared documentation and video tutorials at
+http://next.gs. We have used this tool extensively to simplify our own genome
+projects and believe Genomer may also be useful to others in the field. We will
+illustrate the possible application of Genomer with the following example use
+case.
 
@@
   \item{A Genomer project is organised around a set of already assembled and
-  annotated contigs. For example we assemble our reads into contigs using the
-  A5 pipeline \cite{tritt2012} and and then subsequently annotate them using
-  the Integrated Microbial Genomes resource \cite{markowitz2006}. We then use
-  these contigs as a starting point to build a draft genome sequence in
-  Genomer.}
+  annotated contigs. In our microbial genome projects we assemble our reads
+  into contigs using the A5 pipeline \cite{tritt2012} and then subsequently
+  annotate them using the Integrated Microbial Genomes resource
+  \cite{markowitz2006}. We then use these contigs as a starting point to build
+  a draft genome sequence in Genomer.}
 
   \item{A genome scaffold is written in the scaffolder file format
-  \cite{barton2012}. This specifies the order and orientation of contigs in the
-  scaffold. Any unresolved gap regions can also be specified in this file.
-  Paired-end sequencing or comparison to reference genomes are a source of
-  contig orientation and order. In our genome projects we determine contig
-  order by aligning contigs to a reference genome using nucmer
-  \cite{kurtz2004}. A detailed overview of the scaffolder file format can be
-  found on the Genomer website.}
+  \cite{barton2012}. This specifies the order and orientation of contigs and
+  unresolved gap regions in the scaffold. Paired-read sequencing or comparison
+  to reference genomes can be used as a source of contig orientation and order.
+  As an example, in our own genome projects we determine contig order by
+  aligning contigs to a reference genome using nucmer \cite{kurtz2004}. A
+  detailed overview of the scaffolder file format can be found on the Genomer
+  website.}
 
-  \item{The \verb+genomer view+ command can be used to generate files of the
-  assembled scaffold. These may include the FASTA file of the assembly or a
-  GFF3 file of updated annotations corresponding to the assembled genome
-  sequence. These files can then be used for downstream analyses of the
-  sequence.}
+  \item{The \verb+genomer view+ command is used to generate files of the
+  assembled scaffold, such as the FASTA file of the assembly or a GFF3 file of
+  annotations locations on the assembled genome sequence. These files can then
+  be used for subsequent analyses of the sequence.}
 
@@
   remaining gaps. During our microbe sequencing projects we closed gaps through
-  a combination of Sanger PCR and \emph{in silico} analysis. Changes to the
-  scaffold file automatically propagate updates in the generated files when the
-  \verb+genomer view+ command is rerun.}
+  a combination of Sanger PCR and \emph{in silico} analysis. Genomer allows
+  changes in the scaffold file to automatically propagate to the required
+  changes in the generated files when the \verb+genomer view+ command is
+  rerun.}
 
-  \item{When researchers are satisfied with the status of their draft the
-  required files for submission to GenBank can be generated: assembled FASTA,
-  contig FASTA, annotation tables, and scaffold layout in AGP. Additional
-  options are used to add the required prefix to gene annotation IDs and begin
-  locus ID numbering at the origin of the sequence. Genomer does not produce
-  the .asn file which is submitted to GenBank, but instead the files that may
-  be used as the input to tbl2asn or sequin. }
+  \item{The required files for submission to GenBank may be generated when
+  researchers are satisfied with the status of their draft. These files include
+  the assembled FASTA file, contig FASTA file, annotation table file, and AGP
+  scaffold file. Additional options can be used to add the required prefix to
+  gene annotation IDs and begin locus ID numbering at the origin of the
+  sequence. Genomer does not produce the .asn file which is submitted to
+  GenBank, but instead the files that may be used as the input to tbl2asn or
+  sequin provided by GenBank.}
 
@@
 
-Genomer was written to satisfy our needs for building and finishing a draft
-microbial genomes less than 10MBp in size. Draft sequence assembly takes place
-in memory and therefore building very large sequences in memory may limit
-performance. We point out however that sizes of even 3GBp should fit in the
-memory of a modern laptop or desktop computer.
+Genomer was written to satisfy our needs for building and finishing megabase
+sized draft genomes where assembly of the draft sequence takes place in memory.
+This makes Genomer well-suited for microbial genome projects where
+approximately 99\% of microbial genomes sequenced to date are less than 10MBp.
+The Assembly of very large sequences may however limit performance when the
+size of the sequence exceeds available. We however anticipate that genome sizes
+of even 3GBp will fit in the memory of a modern laptop or desktop computer.
 
-Genomer simplifies the process of generating the annotation table file required
-for submission to GenBank where only gene entries are required. Specifically a
-user need only provide a GFF3 file of annotations and Genomer will generate the
-additional encoded protein features. This however assumes no intron/exon
-structure and therefore users wishing to submit annotations with alternative
-splicing cannot take advantage of this Genomer feature.
+One features of Genomer is the simplification of generating the annotation
+table file required for submission to GenBank. Genomer requires only a GFF3
+file containing gene-type entries be required. Genomer is able to generate the
+additional encoded protein features from these entries. One caveat to this
+however is that the assumption there is no intron/exon structure and therefore
+users wishing to submit annotations with alternative splicing cannot take
+advantage of this Genomer feature.
 

### 2013-04-29 article.tex
Proof read manuscript
Signed-off-by: Michael Barton <mail@michaelbarton.me.uk>

@@
 edited instead of large sequence files. Genomer is written as a command line
-tool and is an effort to make the manual, and increasingly common, process of
-genome scaffolding more robust and reproducible. Extensive documentation and
-video tutorials are available at http://next.gs.
+tool and is an effort to make the manual process of genome scaffolding more
+robust and reproducible. Extensive documentation and video tutorials are
+available at http://next.gs.
 
@@
 
-Manually editing a genome sequence or annotation is a non-trivial problem
-requiring effort on a researcher's part. The large size of FASTA sequence files
-or general feature format (GFF) annotation files makes changing these in a text
+Manually editing a genome sequence or annotation is non-trivial and requires
+effort on a researcher's part. The large size of FASTA sequence files or
+general feature format (GFF) annotation files makes changing these in a text
 editor difficult. Tasks such as organising project files, adding additional
@@
 Genomer is written as a command-line tool for use in shells such as bash or
-zsh. The advantage of command-line tools over GUI-based tools, is that the former
-are simpler to automate and allow the sharing of steps with other researchers
-working on related projects. For instance, during our microbial genome project
-we automated Genomer using GNU Make build files, allowing the finishing steps
-to be repeated automatically using the \verb+make+ command. Genomer may
-therefore be of particular interest to bioinformaticians coordinating in groups
-on the same genome project and those who prefer automated approaches and
-scripting on the command line.
+zsh. The advantage of command-line tools over GUI-based tools, is that the
+former are simpler to automate and allow the sharing of scripts with other
+researchers working on related projects. For instance, during our microbial
+genome project we automated Genomer using GNU Make build files, allowing the
+finishing steps to be repeated automatically using the \verb+make+ command.
+Genomer may therefore be of particular interest to bioinformaticians
+coordinating in groups on the same genome project and those who prefer
+automated approaches and scripting on the command line.
 
@@
 Genomer was written to satisfy our needs for building and finishing megabase
-sized draft genomes where assembly of the draft sequence takes place in memory.
-This makes Genomer well-suited for microbial genome projects where
-approximately 99\% of microbial genomes sequenced to date are less than 10MBp.
-The Assembly of very large sequences may however limit performance when the
-size of the sequence exceeds available. We however anticipate that genome sizes
+sized draft genomes where assembly of the sequence takes place in memory. This
+makes Genomer well-suited for microbial genome projects where approximately
+99\% of microbial genomes sequenced to date are less than 10MBp. The assembly
+of very large sequences may however limit performance when the size of the
+sequence exceeds available RAM space. We however anticipate that genome sizes
 of even 3GBp will fit in the memory of a modern laptop or desktop computer.
@@
 table file required for submission to GenBank. Genomer requires only a GFF3
-file containing gene-type entries be required. Genomer is able to generate the
+file containing gene-type entries. Genomer is then able to generate the
 additional encoded protein features from these entries. One caveat to this

### 2013-05-06 article.tex
Final proof read
Signed-off-by: Michael Barton <mail@michaelbarton.me.uk>

@@
 learning curve in the subsequent computational steps required to process raw
-sequence reads into a database-deposited genome sequence. Here we describe
-``Genomer,'' a tool to simplify the manual tasks of finishing and uploading a
-genome sequence to a database. Genomer can format a genome scaffold into the
-common files required for submission to GenBank. This software also simplifies
+reads into a database-deposited genome sequence. Here we describe ``Genomer,''
+a tool to simplify the manual tasks of finishing and uploading a genome
+sequence to a database. Genomer can format a genome scaffold into the common
+files required for submission to GenBank. This software also simplifies
 updating a genome scaffold by allowing a human-readable YAML format file to be
@@
 $10e^{6}$-$10e^{10}$ short nucleotide reads, which must be assembled into
-larger contigs which are then scaffolded into larger megabase fragments or
-complete chromosomes. These larger DNA sequences are then annotated with
-genomic features such as protein-coding genes. Each of these required steps can
-however produce imperfect results and remains the subject of active research
+larger contigs that are scaffolded into larger megabase fragments or complete
+chromosomes. These larger DNA sequences are then annotated with genomic
+features such as protein-coding genes. However each of these required steps can
+produce imperfect results and remains the subject of active research
 \cite{earl2011,quail2012,beckloff2012}. Therefore manual curation of a genome
@@
 repetition of effort. In contrast open-source projects can lead to a pooling of
-community effort resulting in a higher standard of software and preventing
+community effort resulting in a higher standard of software and prevents
 reinvention of the same code \cite{ince2012}.
@@
 we developed a tool, called ``Genomer,'' to ease and simplify many of the
-common steps in our genome project. While several graphical user interface
-tools are available to work with genome projects, including G-InforBio,
-Genquire, Pile-LineGUI, Artemis and Consed \cite{tanaka2006, wilkinson2002,
-lopez2011, carver2012, gordon1998}. Genomer provides a simple command line
+common steps in our genome project. Several graphical user interface tools are
+available to work with genome projects, including G-InforBio, Genquire,
+Pile-LineGUI, Artemis and Consed \cite{tanaka2006, wilkinson2002, lopez2011,
+carver2012, gordon1998}. Genomer instead provides a simple command line
 interface which allows scriptable automation and reproducible results. We made
@@
   specified. This thereby simplifies the process of re-organising and trimming
-  contigs in draft genome without having to edit large nucleotide sequences
+  contigs in a draft genome without having to edit large nucleotide sequences
   manually.}
@@
   additional PCR sequences can be used to close gaps in the assembly and the
-  location of all annotations be correspondingly updated.}
+  location of all annotations will be correspondingly updated.}
 
@@
   the generation of all these files from the scaffold file and corresponding
-  GFF3 file and removes the effort to produce these manually.}
+  GFF3 file thereby removing the effort to produce these manually.}
 
@@
   This eliminates a common problem in bioinformatics where software may require
-  technical expertise to first compile and install before being used. Install
+  technical expertise to first compile and install before being used. Installed
   plugins may also be locked to specific versions to prevent backwards
@@
   tool around plain text files. This allows for integration with common Linux
-  tools tools such as GNU Make or git thereby allowing reproducible scripting
-  of the genome finishing process.}
+  tools tools such as GNU Make or git allowing reproducible scripting of the
+  genome finishing process.}
 
@@
 projects and believe Genomer may also be useful to others in the field. We will
-illustrate the possible application of Genomer with the following example use
-case.
+illustrate the possible application of Genomer with the following use case.
 
@@
   As an example, in our own genome projects we determine contig order by
-  aligning contigs to a reference genome using nucmer \cite{kurtz2004}. A
-  detailed overview of the scaffolder file format can be found on the Genomer
-  website.}
+  aligning to a reference genome using nucmer \cite{kurtz2004}. A detailed
+  overview of the scaffolder file format can be found on the Genomer website.}
 
@@
   a combination of Sanger PCR and \emph{in silico} analysis. Genomer allows
-  changes in the scaffold file to automatically propagate to the required
-  changes in the generated files when the \verb+genomer view+ command is
-  rerun.}
+  changes in the scaffold file to automatically propagate to the generated
+  files when the \verb+genomer view+ command is rerun.}
 
@@
   sequence. Genomer does not produce the .asn file which is submitted to
-  GenBank, but instead the files that may be used as the input to tbl2asn or
-  sequin provided by GenBank.}
+  GenBank, but instead the files that are used as the input to tbl2asn or
+  sequin tools provided by GenBank.}
 
@@
 video tutorials on the http://next.gs website. This tool has been useful in our
-own genome projects and we believe it will also be useful to other researchers
+own genome projects, and we believe it will also be useful to other researchers
 especially smaller research groups entering the field for the first time.

### 2013-05-18 article.tex
Fix address line for publication
Signed-off-by: Michael Barton <mail@michaelbarton.me.uk>
Remove acknowledgements for publication
Signed-off-by: Michael Barton <mail@michaelbarton.me.uk>

@@
 \\
-\bf{1} Biology Department, The University of Akron, Akron, OH, 44325-3908, USA
+\bf{1} Biology Department, The University of Akron, Akron, OH, USA.
 \\
@@
 
-This work was supported by the National Institute for Health: IDeA Network of
-Biomedical Research Excellence (KY-INBRE) grant (NIH 2P20 RR016481-09) and the
-NIH R15 AREA Program grant (R15GM079775).
-
 %\section*{References}
